---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---
## Load environment
```{r echo=TRUE}
library(data.table)
library(dplyr)
library(lubridate)
library(knitr)
```

## Loading and preprocessing the data
Download data  
```{r echo=TRUE}
Sys.time()
```
[1] "2015-08-12 13:47:26 CDT"

### Unzip
TODO: Change into an if statement.
```{r echo=TRUE}
unzip("activity.zip")
```

### Read
```{r echo=TRUE}
rawdata <- read.csv("activity.csv")
```


### Examine
```{r echo=TRUE}
summary(rawdata) #Use this. It gives the best overview. Don't have to do the NA check below.

```
The data is comprised of 17568 observations of three variable: steps, date, interval.  
There are 61 days; however, only 53 have data.  
The remaining 8 days are populated with NA values; NA values are only in these 8 days.  
#53 of the 61 dates have data. Eight are NA.  
# > length(table(subdate2$sumbydate))  
# [1] 53  
For each day, there are 288 observations (12x24). An observation is taken at 5 minute intervals.

```{r echo=TRUE}
head(rawdata)

```

#### Create data table.
To enable use of dplyr and lubridate functions.
```{r echo=TRUE}
dtdata <- data.table(rawdata)
dtdata
```


## What is mean total number of steps taken per day?
For this part of the assignment, you can ignore the missing values in
the dataset.

1. Make a histogram of the total number of steps taken each day.  
2. Calculate and report the **mean** and **median** total number of steps taken per day.  

### Calculate sum, mean, and median by date.
```{r echo=TRUE}
calcsfordate <- summarize(group_by(dtdata,date),
                         totalfordate = sum(steps),
                         averagefordate = mean(steps),
                         medianfordate = as.double(median(steps, na.rm=TRUE))
                         )
print(calcsfordate[,]) #TODO I want to see all 61 obs
write.table(calcsfordate) #TODO This works but not formatted
kable(calcsfordate) #Why did this print to the console but not to knitr? 
```

### Graph total number of steps taken
I don't really see what this shows. It doesn't seem to match the table of the same variable. It should be ordered by date already, but it doesn't seem to be.

```{r echo=TRUE}
hist(calcsfordate$totalfordate, breaks=61)
```

```{r echo=TRUE}
#total number of steps taken each day
table(calcsfordate$totalfordate) #53 days. Not in date order.

#mean number of steps taken each day
calcsfordate$averagefordate

#median number of steps taken each day
calcsfordate$averagefordate

select(calcsfordate, averagefordate, medianfordate)

View(select(calcsfordate, averagefordate, medianfordate)) # TODO I want to see all 61 obs
```

The NA values are in entire days. So those days are NA.


### Calculate sum, mean, and median by interval.
On the intervals you have to remove NAs since every interval has 8 instances of NA.s
```{r echo=TRUE}

calcsforinterval <- summarize(group_by(dtdata,interval),
                             totalforinterval = sum(steps, na.rm=TRUE),
                             averageforinterval = mean(steps, na.rm=TRUE),
                             medianforinterval = as.double(median(steps, na.rm=TRUE))
                             )

print(calcsforinterval)
```



## What is the average daily activity pattern?
1. Make a time series plot (i.e. `type = "l"`) of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all days (y-axis)

2. Which 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps?
# Answer Eyeballing: 104      835        10927    206.1698113            19

#TODO Use new variable names
```{r echo=TRUE}
intervalmeans <- subinterval3$meanbyinterval
max(intervalmeans) #206.1698 average number of steps
which(dtdata[,subinterval3$meanbyinterval] == max(intervalmeans)) #indices 104
dtdata[104,] #835 This must be wrong. I
subinterval3[104,] #Maybe it's right 835.

```

max(subinterval3$meanbydate)

## Imputing missing values
Note that there are a number of days/intervals where there are missing
values (coded as `NA`). The presence of missing days may introduce
bias into some calculations or summaries of the data.

1. Calculate and report the total number of missing values in the dataset (i.e. the total number of rows with `NA`s)

2. Devise a strategy for filling in all of the missing values in the dataset. The strategy does not need to be sophisticated. For example, you could use the mean/median for that day, or the mean for that 5-minute interval, etc.

3. Create a new dataset that is equal to the original dataset but with the missing data filled in.

4. Make a histogram of the total number of steps taken each day and Calculate and report the **mean** and **median** total number of steps taken per day. Do these values differ from the estimates from the first part of the assignment? What is the impact of imputing missing data on the estimates of the total daily number of steps?

A1: Eight days (columns) have NAs so ALL the rows have NAs. Isn't it more interesting to know which of the columns(dates) have NAs?
288*8= 2304 = Total NAs. This is confirmed in the summary.


summary(dtdata)
countofmissing <- count(is.na(dtda)

bad <- which(is.na(dtdata$steps)) # returns indices
So the answer is 2304?
dtdata[1:20, 1]

A2: Use mean for interval to fill in missing NA




## Are there differences in activity patterns between weekdays and weekends?
For this part the `weekdays()` function may be of some help here. Use
the dataset with the filled-in missing values for this part.

1. Create a new factor variable in the dataset with two levels -- "weekday" and "weekend" indicating whether a given date is a weekday or weekend day.

1. Make a panel plot containing a time series plot (i.e. `type = "l"`) of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all weekday days or weekend days (y-axis). The plot should look something like the following, which was created using **simulated data**:


library(lubridate)

